---
title: "The elliptic package: Weierstrass and Jacobi elliptic functions in R"
output:
  github_document:
    pandoc_args: --webtex
---
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


# elliptic: Weierstrass and Jacobi elliptic functions


[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/elliptic?color=green)](https://cran.r-project.org/package=elliptic)
![](https://cranlogs.r-pkg.org/badges/grand-total/elliptic?color=green)
![](https://cranlogs.r-pkg.org/badges/elliptic?color=green)
![](https://cranlogs.r-pkg.org/badges/last-week/elliptic?color=green)
[![Rdoc](https://www.rdocumentation.org/badges/version/elliptic)](http://www.rdocumentation.org/packages/elliptic)


```{r echo=FALSE, results="hide", message=FALSE}
library("elliptic")
```

[![saythanks](https://img.shields.io/badge/say-thanks-ff69b4.svg)](https://saythanks.io/to/RobinHankin)

----


## Overview

An elliptic function is a meromorphic complex function that is
periodic in two directions.  That is, there exist two nonzero complex
numbers $\omega_1,\omega_2$ with
$\omega_1/\omega_2\in\mathbb{C}\backslash\mathbb{R}$ such that

\[
f(z) = f(z+2\omega_1)=f(z+2\omega_2)
\]

whenever $f(z)$ is defined; note carefully the factors of 2.  There
are two natural ways of presenting elliptic functions: that of
Weierstrass, and that of Jacobi.  Historically, the Jacobi form was
first presented and is the most practically useful, but the
Weierstrass form is more elegant (IMO).

Terminology follows that of Abramowitz and Stegun wherever possible.

# Installation

To install the most recent stable version on CRAN, use
```install.packages()``` at the R prompt:

```
R> install.packages("elliptic")
```


To install the current development version use ```devtools```:

```
R> devtools::install_github("RobinHankin/elliptic")
```

And then to load the package use ```library()```:


```{r,message=FALSE}
library("elliptic")
```

The package comes with an extensive and detailed vignette; type 
```vignette("ellipticpaper")``` at the R commandline.

# The package in use

The Weierstrass elliptic function is evaluated numerically by
```P()```, which takes the half periods $\omega_1,\omega_2$.  Thus


```{r}
z <- 0.3 + 0.2i
omega1 <- 5+1i; omega2 <- 1+7i  # half-periods
f <- function(z){P(z,Omega=c(omega1,omega2))}
c(f(z),f(z + 10+2i), f(z + 2+14i))  # should be equal
```

The elliptic functions can  be visualised using ```view()```:

```{r view_wp}
x <- seq(from=-4, to=4, len=200)
y <- x
z <- outer(x,1i*x, "+")
f <- P(z, c(1+1i,2-3i))
par(pty="s")
view(x,y,f,real.contour=FALSE,drawlabel=FALSE,axes=FALSE,xlab="Re(z)",ylab="Im(z)", main="P(z,1+i,2-3i)")
axis(1,pos = -4)
axis(2,pos = -4)
lines(x=c(-4,4),y=c(4,4))
lines(y=c(-4,4),x=c(4,4))
```


Related functions include $\sigma(\cdot)$ (```sigma()``` and the
$\zeta(\cdot)$ (```zeta()```).


## Jacobi forms

Jacobi's elliptic functions are implemented in the package with their
standard names ```sn()```, ```cn()```, ```dn()``` etc.  For example:


```{r view_sn}
view(x,y,sn(z,m=6),real=FALSE,drawlabel=FALSE,axes=FALSE,xlab="Re(z)",ylab="Im(z)", main="The Jacobi sn() function")
axis(1,pos = -4,at=c(-4,-2,0,2,4))
axis(2,pos = -4,at=c(-4,-2,0,2,4))
lines(x=c(-4,4),y=c(4,4))
lines(y=c(-4,4),x=c(4,4))
```

The Jacobi forms are useful in physics and we can use them to visualise potential flow in a rectangle:

```{r view_rectangle}
n <- 300
K <- K.fun(1/2)  # aspect ratio
f <- function(z){1i*log((z-1.7+3i)*(z-1.7-3i)/(z+1-0.3i)/(z+1+0.3i))} # position of source and sink
x <- seq(from=-K,to=K,len=n)
y <- seq(from=0,to=K,len=n)
z <- outer(x,1i*y,"+")

view(x, y, f(sn(z,m=1/2)), nlevels=44, real.contour=TRUE, drawlabels=FALSE,
     main="Potential flow in a rectangle",axes=FALSE,xlab="",ylab="")
rect(-K,0,K,K,lwd=3)
```


## References

* M Abramowitz and IA Stegun (1965).  _Handbook of Mathematical Functions_.  New York: Dover
* RKS Hankin (2006). "Introducing elliptic, an R package for elliptic and modular functions".  _Journal of Statistical Software_, 15:7
* K Chandrasekharan (1985). _Elliptic functions_.  Springer-Verlag

